---
title: "Model Complete Cases"
author: "Martin Skarzynski"
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(message = FALSE, warnings = FALSE)

```{r libs}
library(here)
library(readr)
library(dplyr)
library(tidyr)
library(survey)
library(purrr)
library(furrr)
plan(multiprocess)
source("kod/get_modelstats.R")

```{r chosen-variables}
#Variables to be included in the model
chsn <- c(
          "PERMTH_INT", #time
          "canc_mort", #event
          "SDPPSU6", #PSU
          "SDPSTRA6", #Stratification
          "WTPFQX6", #Weights
          "HSAITMOR", #Age in months at interview (screener)
          "DMAETHNR", #Ethnicity
          "HSSEX" #Biological sex
      )

```{r remove-variables}
#Variables to be removed from the model
remv <- c(
          "HAN9", #remove age variables
          "HAQ7",
          "HAT29",
          "HAJ0",
          "WTPXRP2", #remove unneed weight variables
          "WTPQRP21",
          "WTPQRP27",
          "WTPQRP43"
          )
path <- 'dat/3-clean-complete-cases.rds'

```{r map-models}
# the modelstats function returns a dataframe
# after taking in the following arguments
#seed,
#n_random_vars,
#datafile_path,
#chosen_vars,
#remove_vars
map_sizes <- function(seed){
map2_dfr(.x = seed,
         .y = seq(10), # number of random variables
         ~get_modelstats(seed=.x,
                         n_random_vars=.y,
                         datafile_path=path,
                         chosen_vars=chsn,
                         remove_vars=remv))
}
#save an object with 1000 models
map_dfr(.x = seq(50), # number of seeds
               .f = ~map_sizes(seed=.x)) %>%
write_rds(here("dat/4-model-first-run.rds"))
```


---
title: "4-model-complete-cases.R"
author: "marskar"
date: "Fri Apr 20 21:14:07 2018"
---
