---
title: Lethal Cancer Risk Factors in NHANES
subtitle: Multidimensional Analysis of <br> Potential Cancer Mortality Risk Factors in the <br> National Health and Nutrition Examination Survey
author:
- Martin Skarzynski
- "Capstone Advisor: Professor Elizabeth Platz"
date: "`r Sys.Date()`"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE,
               warning = FALSE, message = FALSE)
```

```{r libs}
library(readr)
library(here)
library(plotly)
library(knitr)
```

# Goal
1. To explore the exposures measured in the Third National Health and Nutrition Examination Survey (NHANES III) dataset in search of factors associated with cancer mortality and
2. to assess a variable selection method for cancer risk prediction models.

# Data
The core data and corresponding documentation are for the survey interview and examination components are found in four separate data files:
[Lab]

# Explore data from the Third National Health and Nutrition Examination Survey (NHANES III)

<center>
```{r plot1}
plt1 <- read_rds(path = here("plotlyFig1.rds"))
plt1
```
</center>

# Analyze variables in the context of a significance threshold (p < 10^-10^)

<center>
```{r plot2}
plt2 <- read_rds(path = here("volc.rds"))
plt2
```
</center>

# Rank the most frequent highly significant variables

<center>
```{r plot3}
plt3 <- read_rds(path = here("plotlyFig3.rds"))
ax <- list(
  title = ""
  #zeroline = FALSE,
  #showline = FALSE,
  #showticklabels = FALSE,
  #showgrid = FALSE
)
plt3 %>% layout(yaxis = ax)
```
</center>

# Apply domain knowledge to the top variables 

|Name     | Median HR|    n|  Description |
|:--------|:---------|:----|:-------------|
|HSAGEIR  |      1.04| 2016| Age in Years |
|HAQ1     |      1.07| 1415| How would you describe the condition of your natural teeth (excellent, very good, good, fair or poor)? |
|HAT2     |      1.50|  384| In the past month, did you jog or run? |
|HAT16    |      1.67|  256| In the past month, did you lift weights? |
|HAB7     |      0.99|  228| In the past 12 months, how many times were you in a nursing home? |
|DMAETHNR |      1.14|  224| Race/Ethnicity |
|HAK9     |      1.23|  216| How many times per night do you usually get up to urinate? |
|HAP2     |      0.81|  179| Do you use glasses, contacts, or both? |
|HAT10    |      1.43|  119| In the past month, did you do other dancing? |
|HAR1     |      0.63|   96| Have you smoked at least 100 cigarettes during your entire life? |

